### 2022-08-24 main/chapter_searches.tex
some autochecking and tweaks

Simulating events from background and signal hypotheses samples from their
respective priors
\begin{equation}
\pi(x) = \prob{x}{h}
\end{equation}
that correspond to Equation~\ref{eqn:searches_prior} by
$b \leftarrow x$ and $a \leftarrow h$.
Although we previously used the same labels $a$ and $b$, those are only
labels --- the words `prior' and `likelihood' state our intent to index into
the left and right arguments respectively.

@@

This expression integral, or average over the prior, means that evidence $Z$
for model arises from a larger fraction of samples from it hitting large values
of the likelihood function.
In contrast, one might consider a maximum likelihood approach that constrained
an optimization to regions of nonzero prior in each model.
Such optimizations can give a useful upper bound on $Z$, but this example
highlights a flaw since both models support the $\met > 100\,\eV[G]$ bin
and both support the same maximum likelihood of
$L(\varepsilon + 100\,\eV[G]) = 1$.

@@

The infamy of Bayes' rule stems from a fixation on the posterior distributions
from unjustified prior assignments.
Classical works by
Bayes~\cite{bayes1763lii} and
Laplace~\cite{laplace1774stigler} began this focus ---
both assume prior densities that are uniform in chosen coordinates, a practice
that can yield practical results in posterior inferences.
But to report scientific data, such practices are fairly criticized as
arbitrary when they influce the reported results.
Prior assignments are arbitrary.
You are welcome to propose any prediction in any circumstances, and the
scientific response is to test whether or not its predictions are compatible
with our observations of nature.
Our emphasis on the likelihood functions and evidence values that quantify
those predictions avoids the pitfall of seeking uninformative, neutral, or
unbiased priors by embracing priors as active and explicit predictions.
This is the modern, practical approach to Bayesian data analysis~\cite{
mackay2003information,
skilling2004nested,
skilling2006nested,
sivia2006data,
skilling2010foundations
}.

@@

In the words of Skilling~\cite{skilling2008rant},
``we SHOULD use Bayes by assigning subjective priors and inspecting evidence
values.''
This framing of probabilistic data analysis agrees exactly with out standard
practice of histogramming simulations,
as illustrated in Figure~\ref{fig:searches_sig_bkg_prior_likelihood}.
Signal and background models define priors, and their (normalized) bin yields
are the evidence values that we use to compare the models.
Finally corresponding posteriors are the normalized shapes those samples take
within the observed histogram bin, which we do not use to compare the models,
but which could in principle be used for secondary interpretations.

\paragraph{Probability assignments.}
Probability algebra manipulates probabilities, but does not state how they
should be initially assigned.
In general, such probability assignments are free and unconstrained, and might
informed by calibration data or loosely motivated opinions.
But there are examples of unambiguously derived that are commonly used in
High-Energy Physics, notably the Poisson and Gaussian distributions.
To avoid turning this thesis into a textbook, we only sketch the derivations
of these, and refer to real textbooks for their detailed
derivations~\cite{jaynes2003probability}, both of which use the exponential
identity that
\begin{equation}
\label{eqn:searches_exponential}
\exp(x) = \displaystyle \lim_{N \to \infty}
\left(1 - \frac{x}{N}\right)^N
.
\end{equation}

The Large Hadron Collider collides $40$~million proton bunches per second,
but \atlas' trigger system accepts only around $1000$ of
those per second~\cite{atlas2020trigger}.
This means that we have scenario with a large fixed number $N$ of trials
(bunch collisions)
and a small probability of binary acceptance (triggering),
due to a notable collision event.
The summed total number $n$ of accepted events is therefore described by the
Binomial distribution for $n$;
\begin{equation}
\mathrm{Binomial}(n\mid N, p) = p^n (1 - p)^{N - n} \times \binom{N}{n}
,
\end{equation}
which is a product of the probabilities (produce rule) in the sequence of $N$
accept/reject decisions, summed (sum rule) over all of that sequence
permutations that could have led to the total $n$.
With our extremely small trigger rate, we are interested in the limit of small
$p$ and large $N$ which result in a mean acceptance number of $\lambda = Np$.
Using the exponential identity of Equation~\ref{eqn:searches_exponential}, one
can show that this limit tends towards the Poisson
distribution~\cite{jaynes2003probability};
\begin{equation}
\label{eqn:searches_poisson}
\mathrm{Poisson}(n\mid \lambda) = \frac{\lambda^n}{n!}\exp(-\lambda)
,
\end{equation}
which is used ubiquitously in the analysis of our histograms.

The Gaussian distribution has many justifications:
as the unique distribution that is spherically symmetric and factors linearly
into independent dimensions~\cite{
jaynes2003probability,
herschel1850normal,
maxwell1860normal,
muller1959note,
marsaglia1972normal
},
as a practical approximation that permits analysis in linear algebra,
and as the maximum entropy~\cite{PhysRev.106.620} distribution for a
known mean and variance.
This maximum entropy statement means that for a distribution with known mean
and variance in a given coordinates, its best Gaussian approximation matches
those moments.\footnote{%
By `best', this means minimal Kullback-Leibler divergence
\(
H(p\leftarrow q) = \sum_i p_i \log(p_i/q_i)
\)
from the Gaussian approximation $q$ to the target $p$ with known moments.
Such an approximation may not be \emph{good}, but it is best in class.%
}
As a limiting approximation, the Gaussian form arises from products of many
functions, such as combinations of many independent measurements that multiply
their likelihood functions.
At a maximum $\check{x}$, the first derivative of a function vanishes and its
second derivative is negative, after Taylor expanding about its maximum this
product looks like
\begin{equation}
f(x)^N = \displaystyle \lim_{N \to \infty}
\left(
1 - \kappa (x - \check{x})^2
\right)^N,
\end{equation}
which, again by Equation~\ref{eqn:searches_exponential}, goes into the Gaussian
form
\begin{equation}
\label{eqn:searches_gaussian}
\mathrm{Gaussian}(x\mid \mu, \sigma)
\,\mathrm{d}x =
\frac{1}{\sqrt{2\pi \sigma^2}}
\exp\!\left(-\frac{1}{2}\frac{(x - \mu)^2}{\sigma^2}\right)
\mathrm{d}x
\end{equation}
where the prefactor is for normalization to a probability distribution over
$x$, and we include the $\mathrm{d}x$ terms as reminders that it is a density
function in that coordinate.

@@

As written in the caption to Figure~\ref{fig:searches_beans}, there is a
notional reward to the first exactly answer.
You are therefore incentivised to answer a) quickly, before other good guesses,
and b) exactly, with no reward for being close to the true answer.
If you are the first to answer, then it makes sense to write what you think
is the most probable number.
But others following you, seeing your answer, should prefer to guess
differently if they want to win (and they think any other guess is plausibly
correct).

@@

Although alternative ideas (such as minimax, minimizing a maximum
loss~\cite{savage1951review}) can empirically work well, this the mathematical
statement of this theorem states that restricting oneself to Bayes decision
rules cannot deny access to optimal decisions.
Alternatives can at best be equivalent.

This theory does not, however, define how to construct that cost function.
Like the prior, it must come from context.
Indeed the two are intertwined --- decision making machines and organisms
are capable of learning to make good decisions without constructing either
explicitly.
Towards our goal of usefully reporting data, such decision theory has been
applied to argue for ways to quote properties of posterior distributions.
For example, a minimizing the expected absolute error $|x - \check{x}$ leads to
reporting the posterior median, and minimizing the expected square error
$(x - \check{x})^2$ leads to reporting the posterior mean.
Had the rules of our bean counting game rewarded you based on such distances,
these results could have been good guesses. But those weren't the rules.

>>>>

Simulating events from background and signal hypotheses samples from their
respective priors
\begin{equation}
\pi(x) = \prob{x}{h}
\end{equation}
that correspond to Equation~\ref{eqn:searches_prior} by
$b \leftarrow x$ and $a \leftarrow h$.
Although we previously used the same labels $a$ and $b$, those are only
labels --- the words `prior' and `likelihood' state our intent to index into
the left and right arguments, respectively.

@@

This expression as an integral --- an average over the prior, means that
the evidence $Z$ depends on what fraction of samples hits large values of
the likelihood function, and not only on how large that likelihood gets.
In contrast, one could consider a maximum likelihood approach that would
constrain a numerical optimization to regions of nonzero prior in each model.
Such optimizations can give a useful upper bound on $Z$, but this example
highlights a flaw when both models support the $\met > 100\,\eV[G]$ bin.
Then, both support the same maximum likelihood of
$L(\varepsilon + 100\,\eV[G]) = 1$, despite different evidence values.

@@

The infamy of Bayes' rule stems from a fixation on the posterior distributions
from unjustified prior assignments.
Classical works by
Bayes~\cite{bayes1763lii} and
Laplace~\cite{laplace1774stigler} began this focus ---
both assume prior densities that are uniform in chosen coordinates, a practice
that can yield practical results in posterior inferences.
But to report scientific data, such practices are fairly criticized as
arbitrary when they influence the reported results.
Prior assignments are arbitrary.
You are welcome to propose any prediction in any circumstances, and the
scientific response is to test whether or not its predictions are compatible
with our observations of nature.
Our emphasis on the likelihood functions and evidence values that quantify
those predictions avoids the pitfall of seeking uninformative, neutral, or
unbiased priors by embracing priors as active and explicit predictions.
This is the modern, practical approach to Bayesian data analysis~\cite{
mackay2003information,
skilling2004nested,
skilling2006nested,
sivia2006data,
skilling2010foundations
}.

@@

Phrased succinctly by Skilling~\cite{skilling2008rant},
``we SHOULD use Bayes by assigning subjective priors and inspecting evidence
values.''
This framing of probabilistic data analysis agrees exactly with our standard
practice of histogramming simulations,
as illustrated in Figure~\ref{fig:searches_sig_bkg_prior_likelihood}.
Signal and background models define priors, and their (normalized) bin yields
are the evidence values that we use to compare the models.
Finally, corresponding posteriors are the normalized shapes those samples take
within the observed histogram bin, with which we do not compare the models,
but which could in principle be used for secondary interpretations.

\paragraph{Probability assignments.}
Probability algebra manipulates probabilities, but does not state how they
should be initially assigned.
In general, such probability assignments are free and unconstrained, and might
be informed by calibration data or loosely motivated opinions.
But there are examples of unambiguously derived that are commonly used in
High-Energy Physics, notably the Poisson and Gaussian distributions.
To avoid turning this thesis into a textbook, we only sketch the derivations
of these, and refer to real textbooks for their detailed
derivations~\cite{jaynes2003probability}, both of which use the exponential
identity that
\begin{equation}
\label{eqn:searches_exponential}
\exp(x) = \displaystyle \lim_{N \to \infty}
\left(1 - \frac{x}{N}\right)^N
.
\end{equation}

The Large Hadron Collider collides $40$~million proton bunches per second,
but \atlas' trigger system accepts only around $1000$ of
those per second~\cite{atlas2020trigger}.
This means that we have a scenario with a large fixed number $N$ of trials
(bunch collisions)
and a small probability of binary acceptance (triggering),
due to a notable collision event.
The summed total number $n$ of accepted events is therefore described by the
Binomial distribution for $n$;
\begin{equation}
\mathrm{Binomial}(n\mid N, p) = p^n (1 - p)^{N - n} \times \binom{N}{n}
,
\end{equation}
which is a product of the probabilities (produce rule) in the sequence of $N$
accept/reject decisions, summed (sum rule) over all permutations of that sequence
that could have led to the total $n$.
With our tiny trigger rate, we are interested in the limit of small
$p$ and large $N$, which result in a mean acceptance number of $\lambda = Np$.
Using the exponential identity of Equation~\ref{eqn:searches_exponential}, one
can show that this limit tends towards the Poisson
distribution~\cite{jaynes2003probability};
\begin{equation}
\label{eqn:searches_poisson}
\mathrm{Poisson}(n\mid \lambda) = \frac{\lambda^n}{n!}\exp(-\lambda)
,
\end{equation}
which is used ubiquitously in the analysis of our histograms.

The Gaussian distribution has many justifications:
as the unique distribution that is spherically symmetric and factors linearly
into independent dimensions~\cite{
jaynes2003probability,
herschel1850normal,
maxwell1860normal,
muller1959note,
marsaglia1972normal
},
as a practical approximation that permits analysis in linear algebra,
and as the maximum entropy~\cite{PhysRev.106.620} distribution for a
known mean and variance.
This maximum entropy statement means that for a distribution with known mean
and variance in given coordinates, its best Gaussian approximation matches
those moments.\footnote{%
By `best', this means minimal Kullback-Leibler divergence
\(
H(p\leftarrow q) = \sum_i p_i \log(p_i/q_i)
\)
from the Gaussian approximation $q$ to the target $p$ with known moments.
Such an approximation may not be \emph{good}, but it is best in class.%
}
As a limiting approximation, the Gaussian form arises from products of many
functions, such as combinations of many independent measurements that multiply
their likelihood functions.
At a maximum $\check{x}$, the first derivative of a function vanishes and the
second derivative is negative.
Taylor expanding about this maximum to the leading order, the product looks like
\begin{equation}
f(x)^N = \displaystyle \lim_{N \to \infty}
\left(
1 - \kappa (x - \check{x})^2
\right)^N,
\end{equation}
which, again by Equation~\ref{eqn:searches_exponential}, goes into the Gaussian
form
\begin{equation}
\label{eqn:searches_gaussian}
\mathrm{Gaussian}(x\mid \mu, \sigma)
\,\mathrm{d}x =
\frac{1}{\sqrt{2\pi \sigma^2}}
\exp\!\left(-\frac{1}{2}\frac{(x - \mu)^2}{\sigma^2}\right)
\mathrm{d}x
\end{equation}
where the prefactor is for normalization to a probability distribution over
$x$, and we include the $\mathrm{d}x$ terms as reminders that it is a density
function in that coordinate.

@@

As written in the caption to Figure~\ref{fig:searches_beans}, there is a
notional reward to the first exactly answer.
You are therefore incentivized to answer a) quickly, before other good guesses,
and b) exactly, with no reward for being close to the true answer.
If you are the first to answer, then it makes sense to write what you think
is the most probable number.
But others following you, seeing your answer, should prefer to guess
differently if they want to win (and they think any other guess is plausibly
correct).

@@

Although alternative ideas (such as minimax, minimizing a maximum
loss~\cite{savage1951review}) can empirically work well, the mathematical
statement of this theorem states that restricting oneself to Bayes decision
rules cannot deny access to optimal decisions.
Alternatives can at best be equivalent.

This theory does not, however, define how to construct that cost function.
Like the prior, it must come from context.
Indeed, the two are intertwined --- decision-making machines and organisms
are capable of learning to make good decisions without constructing either
explicitly.
Towards our goal of usefully reporting data, such decision theory has been
applied to argue for ways to quote properties of posterior distributions.
For example, a minimizing the expected absolute error $|x - \check{x}$ leads to
reporting the posterior median, and minimizing the expected square error
$(x - \check{x})^2$ leads to reporting the posterior mean.
Had the rules of our bean counting game rewarded you based on such distances,
these results could have been good guesses. But those weren't the rules.
